---
title: 'The Pediatric Template of Brain Perfusion: Cross-validation study of supervised multivariate prediction with SiMLR and RGCCA'
author: "Brian B. Avants et al."
date: "`r Sys.Date()`"
output: html_document
urlcolor: blue
---

Similarity-driven multiview linear reconstruction is perhaps best thought of as a
simultaneous dimensionality reduction and variable selection method.  The process
is based on a mixed representation of the joint low-dimensional space of the modalities.
Like other dimensionality reduction tools, these bases can be used for prediction,
clustering or inference.  RGCCA is a closely related method that performs a multiple CCA
and can be used in the same way as SiMLR.  RGCCA can also incorporate sparseness (SGCCA) but
does not incude graph-based regularization or non-negativity constraints.  Furthermore,
SiMLR and RGCCA use very different optimization approaches in order to find their
solution/feature vectors.

In this example, we use RGCCA and SiMLR to jointly decompose matrices of:

* cortical thickness from T1-weighted neuroimaging;

* fractional anisotropy from diffusion tensor neuroimaging;

* cerebral blood flow from arterial spin labeling neuroimaging;

* a demographic matrix that captures the subjects' age, gender and IQ scores.

We assess the success of these decompositions by testing the ability -- in left out data --
of the neuroimaging embeddings to predict age and IQ.
In this comparison, the "best" method produces both minimal error and maximal correlation
for each of the scores.  The error is measured across all cross-validation runs
such that each subject is used as test data in one of the five folds.
Based on benchmark results, the subject's age should be predicted accurately
to (on average) less than 2 years.  See [the PTBP paper](https://pubmed.ncbi.nlm.nih.gov/25977810/)
for details.

## Load relevant packages

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
bd = path.expand( "~/capsule/data/" )
if ( ! dir.exists( bd ) ) bd = "./data/"
set.seed( 808 )
library( pander )
library( ggplot2 )
library( randomForestExplainer )
library(ANTsR)
library(visreg)
library(randomForest)
```

## Basic Setup

Get masks for each modality, along with matrices wherein the normalized
data is stored.

```{r dataio}
neanatvecs=2
thkmask=antsImageRead( paste(bd,"ptbp_mask_thickness.nii.gz",sep='') )
qth=0.05
mth='BH'
demog=read.csv(paste(bd,"ptbp_summary_demographics.csv",sep='') )
demog=demog[ , 1:19 ]
```

## Generic function to collect and organize data

```{r genericvox}
voxandeanatstudy <- function( demog, imgmat, imgmask,
                              formulabase, formulatest,
                              voi,
                              exclusionThresh,
                              baddata,
                              outprefix,
                              nv )
  {
  imgrowmeans=rowMeans( imgmat )
  wp = rep( FALSE, nrow( demog ) )
  for ( sub in unique( demog$SubID ) ) # get unique subjects
    {
    ww=which( demog$SubID == sub )
    ww=ww[ !is.na( imgrowmeans[ww] )  &
             imgrowmeans[ww] > exclusionThresh ]
    if ( length( ww ) > 0 ) wp[ ww[ 1 ] ] = TRUE
  }
 return( wp )
#  if ( ! all( is.na(baddata) ) ) # FIXME
#   wp[  which(wp==TRUE)[ baddata ] ]=FALSE # bad data
  ilist=matrixToImages( imgmat[wp,], imgmask )
#  for ( i in 1:length(ilist) ) ilist[[i]]=smoothImage( ilist[[i]], 2.0 )
  mydf=data.frame(
                 Sex=demog[wp,]$Sex,
                 AgeAtScan=demog[wp,]$AgeAtScan,
                 PIQ=demog[wp,]$Performance.IQ,
                 VIQ=demog[wp,]$Verbal.IQ,
                 BV=demog[wp,]$BV,
                 Cortex=demog[wp,]$Cortex,
                 LadderCom=demog[wp,]$Teen.Ladder.Community.Score,
                 LadderSES=demog[wp,]$Teen.Ladder.SES.score,
                 Income=demog[wp,]$Income )
  for ( kk in 2:ncol(mydf) ) mydf[,kk]=antsrimpute( mydf[,kk] )
  imgmat=imageListToMatrix( ilist, imgmask )
  locform=formula( paste( "imgmat ~", formulabase ,"+", formulatest ) )
  mdl=lm( locform, data=mydf )
  voxlm=bigLMStats( mdl , 1.e-8 )
  print(paste("begin low",outprefix) )
  lowmat = scale( lowrankRowMatrix( imgmat , 10 ), scale=FALSE )
  print(paste("begin eanat",outprefix) )
  imgeanat=sparseDecom( lowmat , imgmask, nvecs=nv, smooth=0.0,
    sparseness=0.05, cthresh=20000, mycoption=1, its=1 )
  rm( lowmat )
  print(paste("end eanat",outprefix) )
  eproj=abs(  imgeanat$eig  )
  eproj=( eproj/rowSums(eproj) )
  imgproj = imgmat %*% t( eproj  )
  mydf=data.frame( mydf, imgproj )
  print( names(mydf) )
  formbase=formula( paste( "imgproj ~", formulabase ) )
  formtest=formula( paste( "imgproj ~", formulabase ,"+", formulatest ) )
  mdl1=lm( formbase, data=mydf )
  mdl2=lm( formtest, data=mydf )
  eanatlm=bigLMStats( mdl2 , 1.e-6 )
  eseg=eigSeg( mask=imgmask, matrixToImages( imgeanat$eig, imgmask), F  )
  ofn=paste( outprefix, '_eseg.nii.gz', sep='' )
  antsImageWrite( eseg , ofn )
  anv=anova( mdl1, mdl2 )
  return( list(mydf=mydf,
               voxlm=voxlm,
               eanatlm=eanatlm,
               eseg=eseg,
               imgproj=imgproj,
               whichSubjects=wp,
               eanat=imgeanat,
               anv=anv) )
  }
```


# Collect each of the voxel-wise matrix modalities


## Cortical Thickness data

The thickness of the cortex is a well-known maturational index with
rapid changes during adolescence.  Here, we use DiReCT in ANTs to estimate
thickness from T1 data and stored the result, after normalization to a group
template, in a matrix.  The PTBP paper describes this in more detail.
The mask covers template cortex.

```{r thkfun}
############################################
thkmask=antsImageRead( paste(bd,"ptbp_mask_thickness.nii.gz",sep='') )
thkfn=paste(bd,"ptbp_vox_thk.mha",sep='')
thkmat=as.matrix( antsImageRead( thkfn ) )
thkmat[ is.na( thkmat ) ] = 0
baddata=NA
if ( ! exists("wthk")  ) {
  wthk=voxandeanatstudy( demog, thkmat, thkmask,
         formulathkbase, formulathktest, voi=voi,
         exclusionThresh = 0.5, baddata=baddata,
         outprefix='/tmp/Eanat_THK', nv=neanatvecs )
#  qv=p.adjust( ethk$eanatlm$beta.pval[ voi, ], method='BH' )
 # print( min(qv) )
  }
```

## Fractional anisotropy (FA) data

FA is a maturational index that relates to the connective structure in the
brain and is complementary to cortical thickness (computation vs wiring).
As with CT, the FA is normalizes to a template and collected into a matrix
here.  The mask covers template white matter.

```{r fafun}
fafn=paste(bd,"ptbp_vox_fa.mha",sep='')
famask=antsImageRead( paste(bd,"ptbp_mask_fa.nii.gz",sep='') )
famat=as.matrix( antsImageRead( fafn ) )
famat[ is.na( famat ) ] = 0
baddata=c(42,44) # FA
if ( ! exists("wfa")  ) {
  wfa=voxandeanatstudy( demog, famat, famask,
         formulathkbase, formulathktest, voi=voi,
         exclusionThresh = 0.2, baddata=baddata,
         outprefix='/tmp/Eanat_FA', nv=neanatvecs )
 # qv=p.adjust( efa$eanatlm$beta.pval[ voi, ], method='BH' )
#  print( min(qv) )
  }
```

## Cerebral blood flow (CBF) data

In contrast to the prior two measurements, CBF is a functional measurement
related to blood flow or perfusion in the brain.  CBF is more variable
during the day and with a variety of other factors in comparison to structural
measurements.  However, it also shows general trends that relate to maturation.
The matrix, here, covers the cortex.

```{r cbffun}
cbffn=paste(bd,"ptbp_vox_cbf.mha",sep='')
cbfmat=as.matrix( antsImageRead( cbffn ) )
cbfmat[ is.na( cbfmat ) ] = 0
baddata=NA
if ( ! exists("wcbf") ) {
  wcbf=voxandeanatstudy( demog, cbfmat, thkmask,
         formulathkbase, formulathktest, voi=voi,
         exclusionThresh = 45, baddata=baddata,
         outprefix='/tmp/Eanat_CBF', nv=neanatvecs )
#  qv=p.adjust( ecbf$eanatlm$beta.pval[ voi,], method='BH' )
 # print( min(qv) )
  }
```

# Joint modeling of thickness, FA and CBF

## SiMLR

We first collect only subjects that have all 3 modalities.

```{r estimateRank}
if ( ! exists( "energyType" ) ) energyType = 'regression'
haveAllMerge=( wthk & wfa & wcbf )
inmats = list( thkmat[haveAllMerge,], famat[haveAllMerge,], cbfmat[haveAllMerge,] )
demogMerge = demog[ haveAllMerge, ]
demogMerge$Performance.IQ = antsrimpute( demogMerge$Performance.IQ )
demogMerge$Verbal.IQ = antsrimpute( demogMerge$Verbal.IQ )
demogMerge$FullScaleIQ = antsrimpute( demogMerge$FullScaleIQ )
mdlMatrix = model.matrix( lm( BV ~
  AgeAtScan + FullScaleIQ + Verbal.IQ + Performance.IQ + Sex,
  data = demogMerge ))[,-1]
if ( nrow(inmats[[1]]) != nrow( mdlMatrix ) )
  stop(" nrow( nrow(inmats[[1]]) ) != nrow( mdlMatrix ) ")
inmats[[4]] = mdlMatrix
rankData = list( suggestedRank = 3 )  # simlrRankEstimate( 0.90, inmats[[1]], inmats[[2]], inmats[[3]] )
```

In this code chunk, we prepare regularization
matrices (using wrapper functions from ANTsR), define the cross-validation
runs and then call SiMLR and RGCCA on each fold, storing appropriate results.

```{r simlr}
if ( ! exists( "smoms") ) { # regularization matrices based on spatial proximity
  smoms = list()
  spatmat = t( imageDomainToSpatialMatrix( thkmask, thkmask ) )
  smoms[[1]] = knnSmoothingMatrix( spatmat, k = 7^3, sigma = 125 )
  spatmat = t( imageDomainToSpatialMatrix( famask, famask ) )
  smoms[[2]] = knnSmoothingMatrix( spatmat, k = 7^3, sigma = 125 )
  smoms[[3]] = smoms[[1]]
  temp = cor( mdlMatrix )
  temp[ temp < 0.5 ] = 0
  smoms[[4]] = temp
}
# define 5-fold cross-validation groups
if ( ! exists( "mixAlg" ) ) mixAlg = 'svd'
nGroups = 5
cvGroups = rep( c(1:nGroups), nrow( inmats[[1]] ) )[ 1:nrow( inmats[[1]] ) ]
for ( whichGroup in 1:nGroups ) {
    isTraining = rep( TRUE, nrow( inmats[[1]] ) )
    isTraining[ cvGroups == whichGroup ] = FALSE
    inmatsTrain = list()
    for ( j in 1:length( inmats ) ) {
      inmatsTrain[[j]] = inmats[[j]][ isTraining, ]
    }
    # use the max rank which is limited by the smallest matrix
    targetRank = ncol( mdlMatrix ) - 1
    { # rgcca
        rgccaT0 = Sys.time()
        library( RGCCA )
        # tau parameters below correspond to inter-battery factor analysis settings
        # https://cran.r-project.org/web/packages/RGCCA/vignettes/vignette_RGCCA.pdf
        myrgcca = sgcca( # this initializes with SVD
          A = inmatsTrain,
          C = 1 - diag( length( inmatsTrain ) ), # cca-like
          c1 = c(0.2,0.2,0.2,1),  # set to roughly match simlr
          ncomp = c( targetRank, targetRank, targetRank, targetRank ),
          scale = TRUE,
          verbose = TRUE )
        myDimRed = list( v=myrgcca$a )
        pred2 = NULL
        rgccaT1 = Sys.time()
        message( paste( difftime( rgccaT1, rgccaT0, units='mins' ) ) )
#        for ( mm in 1:length( myDimRed$v ) )
 #         antsImageWrite( as.antsImage(  myDimRed$v[[ mm ]] ),
  #          paste0("/results/SiMLR_CV",whichGroup,mixAlg,"_kRGCCA", mm, ".mha") )
      }

      { # simlr
      simlrT0 = Sys.time()
      mysym = simlr(
          voxmats = inmatsTrain,
          smoothingMatrices = smoms,
          iterations = 50,
          sparsenessQuantiles = c( rep(0.9,3), 0.5 ),
          initialUMatrix = targetRank,
          energyType = 'regression',
          mixAlg = mixAlg,
          verbose = 1 )
        simlrT1 = Sys.time()
        message( paste( difftime( simlrT1, simlrT0, units='mins' ) ) )
       # for ( mm in 1:length( mysym$v ) )
       #   antsImageWrite( as.antsImage(  mysym$v[[ mm ]] ),
       #     paste0("/results/SiMLR_CV",whichGroup,mixAlg,"_kReg", mm, ".mha") )

        simlrT2 = Sys.time()
        mysymcca = simlr(
            voxmats = inmatsTrain,
            smoothingMatrices = smoms,
            iterations = 50,
            sparsenessQuantiles = c( rep(0.9,3), 0.5 ),
            initialUMatrix = targetRank,
            energyType = 'cca',
            mixAlg = mixAlg,
            verbose = 1 )
        simlrT3 = Sys.time()
        message( paste( difftime( simlrT3, simlrT2, units='mins' ) ) )
        # use below to write feature matrices to disk
        # for ( mm in 1:length( mysymcca$v ) )
        #  antsImageWrite( as.antsImage(  mysymcca$v[[ mm ]] ),
        #    paste0("/results/SiMLR_CV",whichGroup,mixAlg,"_kCCA", mm, ".mha") )
      }
######
###### assessTestingError
######
    myOPRE = "/results/predDF_"
    mdRGCCA = list( v=myrgcca$a )
    myTestList = list( RGCCA = mdRGCCA , SiMLRReg = mysym , SiMLRCCA = mysymcca )
    nms = names( myTestList )
    myVOIs = c( "AgeAtScan", "Performance.IQ", "Verbal.IQ", "FullScaleIQ" )
    predDF = demogMerge[ , myVOIs]
    colnames( predDF ) = paste0( "RGCCAPred",myVOIs )
    predDemog = cbind( demogMerge, predDF )
    predDF = demogMerge[ , myVOIs]
    colnames( predDF ) = paste0( "SiMLRRegPred",myVOIs )
    predDemog = cbind( predDemog, predDF )
    colnames( predDF ) = paste0( "SiMLRCCAPred",myVOIs )
    predDemog = cbind( predDemog, predDF )
    for ( myvoi in myVOIs ) {
      ct = 1
      for ( myDimRed in myTestList ) {
        myColName = paste0(  nms[ct], "Pred" ,  myvoi )
        predDemog[,myColName] = NA
        pthk = as.matrix( inmats[[1]] %*% ( myDimRed$v[[1]] ) )
        pfa = as.matrix(inmats[[2]] %*% ( myDimRed$v[[2]] ))
        pcbf = as.matrix(inmats[[3]] %*% ( myDimRed$v[[3]] ))
        myproj = as.matrix(inmats[[4]] %*% myDimRed$v[[4]])
        colnames(pthk)=colnames(pfa)=colnames(pcbf)=1:ncol(pcbf)
        demogTrTe = cbind( demogMerge, TH=pthk,  CB=pcbf, FA=pfa, outcome = myproj )
        demogTrTe[ , myvoi] = antsrimpute( demogTrTe[ , myvoi] )
        myform = as.formula( paste( myvoi, " ~  Sex +
          TH.1 + TH.2 + TH.3 + TH.4 + CB.1 + CB.2 + CB.3 + CB.4 + FA.1 + FA.2 + FA.3 + FA.4 "  ) )
        m1 = lm( myform, data = demogTrTe[ isTraining, ] ) # simlr VIQ
        predVAR = predict( m1, newdata = demogTrTe[ !isTraining, ] )
        print( paste( myvoi, nms[ct], "ERROR",
          mean( abs( predVAR - demogTrTe[ !isTraining, myvoi ] ), na.rm=TRUE ),
          "CORR", cor( predVAR, demogTrTe[ !isTraining, myvoi]  ) ) )
        predDemog[ !isTraining,myColName] = predVAR
        ct = ct + 1
        }
      }
    outcsv = paste0( myOPRE, whichGroup, mixAlg, ".csv")
    write.csv( predDemog, outcsv, row.names = FALSE )
    }

```

## Final cross-validated results

The summary results are here.

```{r finalResultsNumbers}
library( ANTsR )
library(ANTsR)
if ( ! exists( "mixAlg" ) ) mixAlg = 'svd'
pdfns = Sys.glob( paste0( myOPRE, "*", mixAlg, ".csv" ) )
presser = read.csv( pdfns[1] )
myVOIs = c( "AgeAtScan", "Performance.IQ", "Verbal.IQ", "FullScaleIQ" )
trueVals = presser[ , myVOIs ]
pCols = 20:31
subber = presser[,pCols]
for ( k in 2:length( pdfns ) ) {
  presser2 = read.csv( pdfns[k] )
  presser2 = presser2[,pCols]
  subber[ !is.na( presser2 ) ] = presser2[ !is.na( presser2 ) ]
  }
plist = list( 1:4, 5:8, 9:12 )
summarizeAll = trueVals
for ( ww in 1:length( plist ) ) {
  subberLoc = subber[,plist[[ww]]]
  summarizeAll = cbind(summarizeAll, subberLoc)
  cat("*****************************\n")
  for ( k in 1:4 ) {
    print( paste( colnames( subberLoc )[k],
      mean(  abs( trueVals[,k] - subberLoc[,k]), na.rm=T ),
      cor( antsrimpute(trueVals[,k]), antsrimpute(subberLoc[,k]) ) ) )
    }
  }
cat("*****************************\n")
write.csv( summarizeAll, paste0( dirname(myOPRE),"/supervised_ptbp_summary_results_", mixAlg, ".csv"), row.names = FALSE )
layout( matrix( 1:3, nrow=1, byrow=TRUE ) )
mylims = range( summarizeAll$AgeAtScan ) + c(-0.5,0.5)
eage = round(mean(abs(summarizeAll$AgeAtScan-summarizeAll$RGCCAPredAgeAtScan))*1000)/1000
plot( summarizeAll$AgeAtScan, summarizeAll$RGCCAPredAgeAtScan,
  main=paste0('RGCCA-scca: ', eage,' years MAE'), xlim=mylims,ylim=mylims, xlab='True Age', ylab='Predicted Age')
eage = round(mean(abs(summarizeAll$AgeAtScan-summarizeAll$SiMLRRegPredAgeAtScan))*1000)/1000
plot( summarizeAll$AgeAtScan, summarizeAll$SiMLRRegPredAgeAtScan,
  main=paste0('SiMLR-Reg-Mix-',mixAlg,': ', eage,' years MAE '), xlim=mylims,ylim=mylims, xlab='True Age', ylab='Predicted Age')
eage = round(mean(abs(summarizeAll$AgeAtScan-summarizeAll$SiMLRCCAPredAgeAtScan))*1000)/1000
plot( summarizeAll$AgeAtScan, summarizeAll$SiMLRCCAPredAgeAtScan,
  main=paste0('SiMLR-CCA-Mix-',mixAlg,': ', eage,' years MAE '), xlim=mylims,ylim=mylims, xlab='True Age', ylab='Predicted Age')
#
# mylims = range( summarizeAll$FullScaleIQ ) + c(-0.5,0.5)
# plot( summarizeAll$FullScaleIQ, summarizeAll$RGCCAPredFullScaleIQ,
#  main='RGCCA-scca: 10.33 MAE', xlim=mylims,ylim=mylims, xlab='True FIQ', ylab='Predicted FIQ')
# plot( summarizeAll$FullScaleIQ, summarizeAll$SiMLRRegPredFullScaleIQ,
#  main='SiMLR-Reg: 9.88 MAE', xlim=mylims,ylim=mylims, xlab='True FIQ', ylab='Predicted FIQ')
# plot( summarizeAll$FullScaleIQ, summarizeAll$SiMLRCCAPredFullScaleIQ,
#  main='SiMLR-CCA: 11.1 MAE', xlim=mylims,ylim=mylims, xlab='True FIQ', ylab='Predicted FIQ')
```


# Discussion

* Reviewed how to set up simlr for imaging studies including regularization
and parameter setting.

* Compared these methods to traditional dimensionality reduction tools for neuroimaging

* We showed strategies for using these modalities together to find meaningful predictors.

* Performed some visualization and interpretation demonstrating the SiMLR outperforms RGCCA's implementation of sparse CCA for brain-based age prediction.
