---
title: "SiMLR vs RGCCA: Cancer multi-omics data"
output: html_document
urlcolor: blue
---


```{r init,echo=FALSE,message=FALSE,warning=FALSE}
set.seed( 98 )
library( ANTsR )
library( RGCCA )
library( measures )
library( survival )
library( survcomp )
library( dplyr )
if ( ! exists( "energyType" ) ) energyType = 'cca'
if ( ! exists( "mixAlg" ) ) mixAlg =  'svd'
### collect data
if ( ! exists( "organ" ) ) organ = 'kidney'
fns = Sys.glob( paste0( "./multiomics/", organ, "/*txt" ) )
if ( dir.exists( "/capsule/data/" ) )
  fns = Sys.glob( paste0( "/capsule/data/", organ, "/*txt" ) )
if ( dir.exists( "/root/capsule/data/" ) )
  fns = Sys.glob( paste0( "/root/capsule/data/", organ, "/*txt" ) )

if ( organ == "Breast" | organ == "breast" ) {
  organ = "Breast"
  fns = Sys.glob( paste0( "./multiomics/", organ, "/*csv" ) )
  if ( dir.exists( "/capsule/data/" ) )
    fns = Sys.glob( paste0( "/capsule/data/", organ, "/*csv" ) )
  if ( dir.exists( "/root/capsule/data/" ) )
    fns = Sys.glob( paste0( "/root/capsule/data/", organ, "/*csv" ) )
}

if ( length( fns ) != 4 ) stop("Files do not exist")
pSel = 0.2
if ( organ == 'gbm' ) {
  datalabel2 = 'KIRC (kidney)'
  datalabel1 = 'GBM (glioblastoma, brain)'
  clin <- read.table( fns[4],stringsAsFactors = FALSE , header=TRUE)
  clin$PatientID = toupper( clin$PatientID )
  m1 = t( read.table( fns[1], header=T ) )
  m2 = t( read.table( fns[2], header=T ) )
  m3 = t( read.table( fns[3], header=T ) )
  m1nms = rownames( m1 )
  m2nms = rownames( m2 )
  m3nms = rownames( m3 )
  unms = intersect( m1nms, m2nms )
  unms = intersect( unms, m3nms )
  clinFilt = clin[ clin$PatientID %in% substr(unms,0,12), ]
  clinFilt = clinFilt[ order( clinFilt$PatientID ), ]
  m1Filt = m1[ m1nms %in% unms, ]
  m2Filt = m2[ m2nms %in% unms, ]
  m3Filt = m3[ m3nms %in% unms, ]
  m1Filt = m1Filt[ order( rownames( m1Filt ) ), ]
  m2Filt = m2Filt[ order( rownames( m2Filt ) ), ]
  m3Filt = m3Filt[ order( rownames( m3Filt ) ), ]
  m1nms = substr(rownames( m1Filt ),0,12)
  m2nms = substr(rownames( m2Filt ),0,12)
  m3nms = substr(rownames( m3Filt ),0,12)
}
if ( organ == 'kidney' ) {
  datalabel1 = 'KIRC (kidney)'
  datalabel2 = 'GBM (glioblastoma, brain)'
  clin <- read.table( fns[4],stringsAsFactors = FALSE , header=TRUE)
  m1 = t( read.table( fns[1], header=T ) )
  m2 = t( read.table( fns[2], header=T ) )
  m3 = t( read.table( fns[3], header=T ) )
  m1nms = rownames( m1 )
  m1nms = gsub(  "[.]", "-" , m1nms )
  m2nms = rownames( m2 )
  m2nms = gsub(  "[.]", "-" , m2nms )
  m3nms = rownames( m3 )
  m3nms = gsub(  "[.]", "-" , m3nms )
  unms = intersect( m1nms, m2nms )
  unms = intersect( unms, m3nms )
  clinFilt = clin[ clin$PatientID %in% unms, ]
  clinFilt = clinFilt[ order( clinFilt$PatientID ), ]
  m1Filt = m1[ m1nms %in% unms, ]
  m2Filt = m2[ m2nms %in% unms, ]
  m3Filt = m3[ m3nms %in% unms, ]
  m1Filt = m1Filt[ order( rownames( m1Filt ) ), ]
  m2Filt = m2Filt[ order( rownames( m2Filt ) ), ]
  m3Filt = m3Filt[ order( rownames( m3Filt ) ), ]
  m1nms = rownames( m1Filt )
  m1nms = gsub(  "[.]", "-" , m1nms )
  m2nms = rownames( m2Filt )
  m2nms = gsub(  "[.]", "-" , m2nms )
  m3nms = rownames( m3Filt )
  m3nms = gsub(  "[.]", "-" , m3nms )
  }
if ( organ == "Breast" ) {
  datalabel2 = 'KIRC (kidney)'
  datalabel1 = 'BIC (breast invasive carcinoma)'
  clinFilt <- read.csv( fns[4] ) %>% distinct()
  clinFilt = clinFilt[ order( clinFilt$PatientID ), ]
  clinFilt$PatientID = toupper( clinFilt$PatientID )

  m1Filt = read.csv( fns[1] ) %>% distinct()
  m1Filt$PatientID=substr(m1Filt$PatientID,0,12)
  m1Filt = m1Filt[ order( m1Filt$PatientID ), ]
  # get first occurrence of each subject
  m1Filt <- m1Filt[match(unique(m1Filt$PatientID), m1Filt$PatientID),]

  m2Filt = read.csv( fns[2] ) %>% distinct()
  m2Filt$PatientID=substr(m2Filt$PatientID,0,12)
  m2Filt = m2Filt[ order( m2Filt$PatientID ), ]
  m2Filt <- m2Filt[match(unique(m2Filt$PatientID), m2Filt$PatientID),]

  m3Filt = read.csv( fns[3] ) %>% distinct()
  m3Filt$PatientID=substr(m3Filt$PatientID,0,12)
  m3Filt = m3Filt[ order( m3Filt$PatientID ), ]
  m3Filt <- m3Filt[match(unique(m3Filt$PatientID), m3Filt$PatientID),]

  usubs = intersect( m1Filt$PatientID, m2Filt$PatientID )
  usubs = intersect( usubs, m3Filt$PatientID )
  usubs = intersect( usubs, clinFilt$PatientID )
  clinFilt = clinFilt[ clinFilt$PatientID %in% usubs, ]
  m1Filt = m1Filt[ m1Filt$PatientID %in% usubs, ]
  m2Filt = m2Filt[ m2Filt$PatientID %in% usubs, ]
  m3Filt = m3Filt[ m3Filt$PatientID %in% usubs, ]
  pSel = 0.2
  m1nms = m1Filt$PatientID
  m2nms = m2Filt$PatientID
  m3nms = m3Filt$PatientID
  # remove pid for consistency with other datasets
  m1Filt = m1Filt[,-1]
  m2Filt = m2Filt[,-1]
  m3Filt = m3Filt[,-1]
}
stopifnot( all.equal(  m1nms, m2nms ) )
stopifnot( all.equal(  m1nms, m3nms ) )
stopifnot( all.equal(  m1nms, clinFilt$PatientID ) )
if ( ! organ %in% c("Breast","kidney", "gbm" ) ) stop("Need to choose organ = gbm, Breast or kidney")
myn = nrow( m1Filt )
myps = c(  ncol(m1Filt), ncol(m2Filt), ncol(m3Filt) )
```

## Multi-omics prediction of cancer survival

We compare two dimensionality reduction methods with respect to
cancer survival prediction based on multi-omics data.  The biological data
includes $n=$ `r myn` subjects with:

* Gene expression: with $p=$ `r myps[1]` predictors; see [Goodwin, et. al. for background](https://pubmed.ncbi.nlm.nih.gov/27184599/)

* Methylomics/DNA methylation: with $p=$ `r myps[2]` predictors; see ["Profiling genome-wide DNA methylation"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4926291/) for background;

* Transcriptomics/micro RNA expression: with $p=$ `r myps[3]` predictors; see ["RNA sequencing: advances, challenges and opportunities" for background](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6202627/);

and is complemented by survival data (number of days since diagnosis
  and whether or not death has occurred at that time).

The hypothesis is that this collection of measurements will improve
our ability to predict patient-specific outcomes.  However, these data are fairly
high-dimensional.  As such, targeted dimensionality reduction is needed to
overcome the $p >> n$ problem (where $p$, here, refers to predictors) in order
to allow low-dimensional versions of these predictors
(i.e. embeddings) to be used in a classical regression context.

The statistical model is a Cox proportional hazards regression model implemented
in the `coxph` function in the `survival` package.  We evaluate concordance in
test data via the `survcomp` package.  Concordance is similar to a rank
correlation method and is used to assess agreement of the predicted
outcomes with true outcomes.  Its value under the null hypothesis of no
predictive value is 0.5.  Values greater than roughly 0.6 show some evidence of
predictive power.

Data is the `r datalabel1` set from [the multi-omic benchmark collection](http://acgt.cs.tau.ac.il/multi_omic_benchmark/download.html).
The published paper associated with this benchmark is [here](https://pubmed.ncbi.nlm.nih.gov/30295871/).
We chose the `r organ` data as it allowed a train-test
split with sufficient variability in both train and test groups.  These data
are collected from The Cancer Genome Atlas (TCGA).

The benchmark paper above showed that
"with respect to survival, MCCA had the total best prognostic value" where
MCCA refers to mulitiple canonical correlation analysis ( pairwise CCA across
  all pairs ).
Thus, the comparison -- done here -- between RGCCA and SiMLR with a CCA-like
objective function is pertinent. Nevertheless, these approaches should not be
considered as the best strategy given that, according to the benchmark
evaluation study, single-omic analysis did nearly as well.
The authors find that "analysis of multi-omics data does not consistently
provide better prognostic value and clinical significance compared to analysis
of single-omic data alone, especially when different single-omics are used
for each cancer types".


The study design is fairly simple.  We first split the data into 80\% training
and 20\% testing sets.  Subsequently, for both RGCCA and SiMLR, we perform:

* in training data, we perform supervised dimensionality
reduction where the 'omics data is jointly reduced
with both death and survival time acting as a fourth matrix (we subsample 95\%
  of the data to do this which allows repeated experiments to be performed);

* training a Cox model with the low-dimensional bases derived from the 'omics
data in the previous step;

* prediction of the outcome in test data and evaluation of prediction
accuracy with the concordance metric.

We repeat the above experiments over $k$ resamplings of the data in order to
gain an estimate of the difference in performance between SiMLR and RGCCA with
different input data.
We also test at two different sparseness levels: moderately sparse (0.5 for both), highly
sparse (0.75 for SiMLR and 0.25 for RGCCA) and low sparseness (0.25, 0.75) where these values impact
the sparseness of the derived feature vectors.

In this evaluation, SiMLR shows a small advantage over RGCCA in terms of
predictive performance as measured by concordance in test data.  Furthermore,
in contrast to RGCCA, SiMLR's feature vectors are not only sparse but also
smooth and unsigned (non-negative) which aids interpretation and may
prevent overfitting, thus improving generalization.  In the example
code below, graph-based regularization parameters are selected to include
roughly 2.5\% of the predictors in each predictor 'omics matrix
(see the call to the `regularizeSimlr` function).  As such, regularization is
present but neither overwhelming nor optimized for this data.  I.e.
this value was chosen based on the desire for a small amount of denoising in
the solution space.  Neither method was optimized for this
problem in terms of data selection, parameter or pre-processing choices.
As such, this acts as a fairly unbiased comparison of these tools.
*To provide further evidence of this, the script can easily be modified to
run on the **`r datalabel2`** data from the benchmark site above*.  In our
experiments, similar trends are found in this related dataset in a different
biological system.

```{r}
myresults = data.frame( simCox = NA, rgCox = NA, sp = NA )
ct = 1
if ( ! exists( "sparSearch" ) ) sparSearch = c( 0.25, 0.50, 0.75 )
nComponents = 2
nBootStrapRuns = 25
for ( sp in sparSearch ) {
  it = 1
  for ( it in it:nBootStrapRuns ) {
    isTrain = rep( TRUE, nrow( m1Filt ) )
    clinFilt[,"Death"] = antsrimpute( clinFilt[,"Death"], FUN=median )
    balancedTest = caret::createDataPartition(  clinFilt[,"Death"],  p=pSel )$Resample1
    isTrain[ balancedTest ] = FALSE
    trainer = list( m1Filt[isTrain,], m2Filt[isTrain, ], m3Filt[isTrain,], clinFilt[isTrain,2:3] )
    tester = list( m1Filt[!isTrain,], m2Filt[!isTrain, ], m3Filt[!isTrain,], clinFilt[!isTrain,2:3] )
    kidney = list( data.train = trainer, data.test = tester )
    fixMat <- function( x ) {
      vv = apply( x, FUN=var, MARGIN=2 )
      ww = which( vv == 0 )
      for ( j in ww  ) {
        x[,ww] = rnorm( nrow(x) )
      }
      scale(x,T,T)
    }
    typTe = kidney$data.test[[4]]
    typTr = kidney$data.train[[4]]
    nnn = 1:nrow( kidney$data.train[[1]] )
    myresults[ct,"sp"] = sp
    trainData = list()
    for ( k in 1:3 )
      trainData[[k]] = fixMat( kidney$data.train[[k]] )
    trainData[[4]] = data.matrix( antsrimpute( kidney$data.train[[4]], median ) )
    # parameters were set s.t. a small amount of gaussian smoothing results
    mysig = 3.5
    if ( ! exists( "regs" ) )
      regs = regularizeSimlr( trainData, fraction=0.025,sigma=c(rep(mysig,3),1))
    # below will give a warning about number of components but ok to ignore
    spn = 1.0 - sp
    sparVals = c( rep( sp, 3 ), 0.0 ) # these lead to roughly similar sparseness
    sparVals2 = c(spn,spn,spn, 1 ) # across both SiMLR and RGCCA

    simresult = simlr(
      trainData,
      regs,
      iterations = 100,
      sparsenessQuantiles = sparVals,
      initialUMatrix = nComponents,
      energyType = energyType,
      mixAlg = mixAlg,
      verbose = T
      )
  #############################################
    proj1 = data.matrix( trainData[[1]] ) %*% simresult$v[[1]]
    proj2 = data.matrix( trainData[[2]] ) %*% simresult$v[[2]]
    proj3 = data.matrix( trainData[[3]] ) %*% simresult$v[[3]]
    trdf = data.frame( proj1=proj1, proj2=proj2, proj3=proj3 )
    proj1 = antsrimpute(data.matrix( kidney$data.test[[1]] ) %*% simresult$v[[1]])
    proj2 = antsrimpute(data.matrix( kidney$data.test[[2]] ) %*% simresult$v[[2]])
    proj3 = antsrimpute(data.matrix( kidney$data.test[[3]] ) %*% simresult$v[[3]])
    tedf = data.frame( proj1=proj1, proj2=proj2 , proj3=proj3 )
    trdf$Survival = antsrimpute( typTr[,1] )
    trdf$Death = antsrimpute(typTr[,2], FUN=median )
    tedf$Survival = antsrimpute(typTe[,1], FUN=median )
    tedf$Death = antsrimpute(typTe[,2], FUN=median )
    coxFitSim <- coxph(Surv(time = Survival , event = Death) ~ . , data = trdf )
    pred_validation = predict( coxFitSim, newdata = tedf )
    # higher values are better - like a rank correlation
    cindex_validation = survcomp::concordance.index(
      pred_validation,
      surv.time  = tedf$Survival,
      surv.event = tedf$Death, method = "noether" )$c.index
    myresults[ct, "simCox"] = cindex_validation
    cindex_validation

    # now do RGCCA
    myrgcca = sgcca( # this initializes with SVD
              A = trainData,
              C = 1 - diag( length( trainData ) ), # cca-like
              c1 = sparVals2,  # set to roughly match simlr
              ncomp = c( nComponents, nComponents, nComponents, 2 ),
              scale = TRUE,
              verbose = FALSE )

    proj1 = data.matrix(trainData[[1]]) %*% myrgcca$a[[1]]
    proj2 = data.matrix(trainData[[2]]) %*% myrgcca$a[[2]]
    proj3 = data.matrix(trainData[[3]]) %*% myrgcca$a[[3]]
    trdf = data.frame( proj1=proj1, proj2=proj2, proj3=proj3 )
    proj1 = antsrimpute(data.matrix(kidney$data.test[[1]]) %*% myrgcca$a[[1]])
    proj2 = antsrimpute(data.matrix(kidney$data.test[[2]]) %*% myrgcca$a[[2]])
    proj3 = antsrimpute(data.matrix(kidney$data.test[[3]]) %*% myrgcca$a[[3]])
    tedf = data.frame( proj1=proj1, proj2=proj2 , proj3=proj3 )
    trdf$Survival = antsrimpute( typTr[,1] )
    trdf$Death = antsrimpute(typTr[,2], FUN=median )
    tedf$Survival = antsrimpute(typTe[,1], FUN=median )
    tedf$Death = antsrimpute(typTe[,2], FUN=median )
    coxFit <- coxph(Surv(time = Survival , event = Death) ~ . , data = trdf )
    pred_validation = predict( coxFit, newdata = tedf )
    cindex_validation = survcomp::concordance.index(pred_validation,
      surv.time  = tedf$Survival,
      surv.event = tedf$Death, method = "noether")$c.index
    myresults[ct,"rgCox"] = cindex_validation
    print( myresults[ct,] )
    print( paste(ct, ":", myresults[ct, "simCox"], myresults[ct, "rgCox"]))
    ct = ct + 1
    }
  message("Example CoxFit for SiMLR")
  print( summary( coxFitSim ) )
  message( paste( "Compare concordance indices", sp ) )
  losel = myresults$sp == sp
  losel = losel &
    !is.na( myresults[,  "simCox"] ) &
    !is.na( myresults[,  "rgCox"] )
  if ( sum(losel) > 2 )
  if ( all( ! is.na(myresults[losel, "simCox"] ) ) &
       all( ! is.na(myresults[losel,  "rgCox"] ) ) & ct > 3 ) {
    print( t.test( myresults[losel,"simCox"], myresults[losel,"rgCox"], paired=T ) )
    }
  print( colMeans( myresults[losel,] , na.rm=T ) )
  mylims = range( myresults[losel,1:2] )+c(-0.05,0.05)
  plot( myresults[losel,"rgCox"], myresults[losel,"simCox"],
        main=paste( 'Concordance: SiMLR vs RGCCA @ spar:', sp ),
        ylab=paste0( 'SiMLR Concordance: ', energyType , ' + ', mixAlg ),
        xlab='SGCCA Concordance', ylim=mylims, xlim=mylims,
        cex=2, cex.lab=1.5, cex.main=1.5, cex.axis=1.3 )
  referenceLine = sort(runif(50,mylims[1],mylims[2]))
  mypt6 = rep( 0.6, length( referenceLine ) )
  points( referenceLine, referenceLine, type='l', col='blue', lty=3 )
  points( referenceLine, mypt6, type='l', col='red', lty=4 )
  points( mypt6, referenceLine, type='l', col='lavenderblush4', lty=5 )
  }
##############################
```

Summarize results.

```{r}
knitr::kable( myresults )
```

Better histogram.


```{r histout}
library(rtemis)
temp=myresults[myresults$sp==0.25,1:2]
colnames(temp) = c('SiMLR','SGCCA')
mplot3.x( na.omit(temp), main='Survival Prediction, spar = 0.25',  xlab='Concordance' )
library(rtemis)
temp=myresults[myresults$sp==0.5,1:2]
colnames(temp) = c('SiMLR','SGCCA')
mplot3.x( na.omit(temp), main='Survival Prediction, spar = 0.5',  xlab='Concordance' )
library(rtemis)
temp=myresults[myresults$sp==0.75,1:2]
colnames(temp) = c('SiMLR','SGCCA')
mplot3.x( na.omit(temp), main='Survival Prediction, spar = 0.75',  xlab='Concordance' )
```


Parameters: mixAlg `r mixAlg`, nComponents `r nComponents` and energyType `r energyType`.
