---
title: "SiMLR: BGLR data"
output: html_document
---

We predict BMI from genotype in a train-test design using SiMLR for
supervised multi-view dimensionality reduction and random forest as
a prediction engine.  We compare to RGCCA with parameters that are
set to closely resemble those chosen for SiMLR.

see [https://cran.r-project.org/web/packages/BGLR/BGLR.pdf](https://cran.r-project.org/web/packages/BGLR/BGLR.pdf) and [https://cran.r-project.org/web/packages/BGLR/vignettes/BGLR-extdoc.pdf](https://cran.r-project.org/web/packages/BGLR/vignettes/BGLR-extdoc.pdf)

Collect the packages we need for this experiment.

```{r setup}
set.seed( 1011 )
library( randomForestExplainer )
library( rtemis )
library( ANTsR )
library( ggplot2 )
library( gaston ) # for LD and LD.plot
if ( ! exists( "energyType" ) ) energyType = 'regression'
if ( ! exists( "constraint" ) ) constraint = 'Grassmann'
rf<-usePkg('randomForest')
bg<-usePkg('BGLR')
mlFun <- randomForest
# mlFun <- lm
```

BGLR provides the example data which is 1,814 mice with 10,346 SNPs each.
We split the data ( using `caret` ) into 80\% training and 20\% testing samples.

```{r data}
data(mice)
snps<-mice.X
if ( ! exists( "myld" ) )
  myld = LD(as.bed.matrix( snps ), lim=c(1,ncol(snps)) ) # expensive to compute
nfolds<-8
train <- caret::createDataPartition( mice.pheno[,"Obesity.BMI"], p=0.8)$Resample1
```


We assemble the phenotype matrix from the relationship between various physical
parameters of the mice.  This resembles the original CCA example given
by Hotelling wherein he proposed investigating "individual differences in mental and physical
traits" through "a study of the relations between sets of correlated variates."
We display the first 200 rows of this design matrix with a heatmap.

```{r bmi}
nms = c( "Obesity.BMI"      ,  "Obesity.BodyLength", "Obesity.EndNormalBW", "GENDER"  )
numericalpheno<-data.matrix( mice.pheno[ , nms ] )
numericalpheno[,"GENDER"] = numericalpheno[,"GENDER"] - 1
temp = data.frame( vox = rnorm( length(train)), mice.pheno[train,nms ] )
temp = lm( vox ~ ( Obesity.BMI  + Obesity.EndNormalBW + Obesity.BodyLength ) * GENDER, data=temp )
mdlMat = (model.matrix( temp ))[,-1]
snpsf = snps # quantifySNPs( snps ) # convert to frequency
inmats = list(
  as.matrix(snpsf[train,]),
  mdlMat )
BMINAME="Obesity.BMI"
# pheatmap::pheatmap( snps[1:200,1:40], cluster_rows=F, cluster_cols=F, show_rownames = F )
pheatmap::pheatmap( mdlMat[1:200,], cluster_rows=F, cluster_cols=F, show_rownames = F )
```

Now we demonstrate using `simlr` (the code) to perform a supervised
multi-view regression between SNPs and the phenotype variables.
First, we produce regularization matrices.  We use linkage
disequilibrium for SNPs and a thresholded correlation matrix for phenotype.

```{r sym}
regs = list()
# demonstrate use of regularization of the high-dimensional matrix
regs[[1]] = as.matrix( myld ) # diag(ncol(inmats[[1]]))
corThresh = c(0.80,0.50)
regs[[1]][regs[[1]]<corThresh[1]]=0
regs[[1]] = as( regs[[1]], "sparseMatrix" )
# regs = regularizeSimlr( inmats, c( 55, 1 ), sigma = c( 10, 10 ) )
regs[[2]] = cor(inmats[[2]])
regs[[2]][regs[[2]]<corThresh[2]]=0
regs[[2]] = as( regs[[2]], "sparseMatrix" )
# library( plot.matrix )
# pheatmap::pheatmap( regs[[2]], cluster_rows=F, cluster_cols=F, show_rownames = F, main='Regularization matrix for supervising variables related to BMI and gender.')
# pheatmap::pheatmap( regs[[1]][1:250,1:250], main='Corner of regularization matrix for SNPs.')
```

Next, we run `simlr` with several starting points and take the best result.
Like $k$-means, `simlr` has sensitivity to initialization so it is prudent to
use multi-start methods (also standard in $k$-means).  We use explained
variance to determine the best result but one could also use the best energy
obtained.  The algorithm will stop before the max iterations if it detects convergence.

```{r runsym}
bestVarx = 0
nEmbed = 2
mysq = c( 0.80, 0.5 ) # very sparse for demonstration purposes
mixAlg = 'ica' 
initu = initializeSimlr( inmats, nEmbed, jointReduction = T, uAlgorithm = mixAlg )
if ( ! exists("sResult")) {
  for ( k in c(1:6) ) { # take best result over multiple initial starting points
    temp = simlr(
      inmats,
      regs,
      iterations = 100,
      sparsenessQuantiles = mysq,
      initialUMatrix = initu,
      energyType = energyType,
      expBeta = 0.99,
      constraint = constraint,
      mixAlg = mixAlg,
      verbose = TRUE,
      randomSeed = k )
    svx = predictSimlr( inmats, temp )
    print( svx$varx )
    if ( sum( svx$varx ) > bestVarx ) {
      sResult = temp
      bestVarx = sum( svx$varx )
      }
    }
}

# ||<o>||<o>||<o>||<o>||<o>||<o>||<o>||<o>||<o>||
```

Demonstrate post-hoc descriptive statistics and prediction of BMI
from SNPs with the selected prediction model.

```{r postsym}
# layout( matrix( 1:4, nrow=1 ))
plot( ts( sResult$totalEnergy ) )
straindf<-data.frame( bmi=numericalpheno[ train,BMINAME] , G=numericalpheno[ train,"GENDER"],
   snpse=as.matrix( snpsf[train, ] ) %*% as.matrix( sResult$v[[1]] )/norm( sResult$v[[1]],"F") )
stestdf <-data.frame( bmi=numericalpheno[-train,BMINAME] , G=numericalpheno[ -train,"GENDER"],
   snpse=as.matrix( snpsf[-train,] ) %*% as.matrix( sResult$v[[1]] )/norm( sResult$v[[1]],"F")  )
myrf<-mlFun( bmi ~ . , data=straindf[,], localImp = TRUE )
predictedBMI <- predict( myrf, newdata=stestdf )
print( cor.test( predictedBMI, stestdf$bmi ) )
print(paste( "Err",  mean( abs( predictedBMI-stestdf$bmi ) ) ) )
rtemis::mplot3_xy(stestdf$bmi, predictedBMI, fit = "lm",main=paste('Predicted vs real BMI'))
importance_frame <- measure_importance( myrf )
plot_multi_way_importance( importance_frame, size_measure = "p_value" )
#  x_measure = "mse_increase",
#  y_measure = "no_of_nodes",
```

Show LD for the selected SNPs.

```{r myLD}

wvec = abs( sResult$v[[1]][,ncol(sResult$v[[1]])] )
wvecTh = quantile( wvec, 0.9995 )
selector = wvec > wvecTh
mySelected = rownames( sResult$v[[1]] )[ selector ]
www = which( selector )
subsnps = snps[,www]
myldSel = LD(as.bed.matrix( subsnps ), lim=c(1,ncol(subsnps)) ) # expensive to compute
# pdf( "~/code/navy/PING/myldsel.pdf", width=6, height=6 )
LD.plot( myldSel )
# dev.off()


```


Below, we run a comparison to RGCCA where we instantiate a version of this
powerful tool that closely resembles the setup we elected for SiMLR.  We
use a CCA design matrix and relatable sparseness parameters.  Results look
very close to that of SiMLR.

```{r rgcca,eval=TRUE}
# Feb 2024 - RGCCA has changed and seems to not work as reliably in the past - at least with these settings.
# users may try to fix this on their own - or use the containerized version associated with the paper.
message("use RGCCA to compute the same type of representation")
# see the RGCCA package vignette for more information on this
library( RGCCA )
# tau parameters below correspond to inter-battery factor analysis settings
# https://cran.r-project.org/web/packages/RGCCA/vignettes/vignette_RGCCA.pdf
# cca.with.rgcca = sgcca( # this initializes with SVD
#  A = inmats,
#  C = matrix(c(0, 1, 1, 0), 2, 2), # cca
#  c1 = 1.0 - mysq,  # set to match simlr
  # tau = c( 1, 1 ), # for rgcca only
#  ncomp = c( nEmbed, nEmbed ), scale = TRUE, verbose = TRUE )
if ( ! exists("cca.with.rgcca"))
  cca.with.rgcca = rgcca( # this initializes with SVD
    A = inmats,
    C = matrix(c(0, 1, 1, 0), 2, 2), # cca
    sparsity = 1.0 - mysq,  # set to match simlr
    # tau = c( 1, 1 ), # for rgcca only
    init = "svd", # svd blew up for unknown reasons
    method='sgcca',
    ncomp = c( nEmbed, nEmbed ), scale = TRUE, verbose = TRUE )
wsol = cca.with.rgcca
straindf<-data.frame( bmi=numericalpheno[ train,BMINAME], G=numericalpheno[ train,"GENDER"] ,
   snpse=as.matrix( snpsf[train, ] ) %*% as.matrix( wsol$a[[1]] ) )
stestdf <-data.frame( bmi=numericalpheno[-train,BMINAME], G=numericalpheno[ -train,"GENDER"] ,
   snpse=as.matrix( snpsf[-train,] ) %*% as.matrix( wsol$a[[1]] ) )
myrf<-mlFun( bmi ~ . , data=straindf[,], localImp = TRUE )
spreddf <- predict( myrf, newdata=stestdf )
rtemis::mplot3_xy(stestdf$bmi, spreddf, fit = "lm",main=paste('Predicted vs real BMI: SGCCA'))
print( cor.test( spreddf, stestdf$bmi ) )
print(paste("RGCCA Err",  mean(abs(spreddf-stestdf$bmi))  ) )
# importance_frame <- measure_importance( myrf )# 
# plot_multi_way_importance( importance_frame, size_measure = "p_value" )
# plot_multi_way_importance( importance_frame, x_measure = "node_purity_increase",
# y_measure = "no_of_nodes", size_measure = "p_value" )
```
