---
title: 'SiMLR Simulation Study: Demonstrate basic assumptions for SiMLR and compare to SVD'
author: "Brian B. Avants et al."
date: "`r Sys.Date()`"
output: html_document
---

Decode known latent signal distributed across 3 matrices with SiMLR and SVD.
We perform permutation testing on the relationship to determine
empirical distribution (significance) and compare with SVD.

The key metrics will compare the ability to predict the known latent signal
in testing data based on the embeddings derived in an unsupervised dimensionality
reduction step.  This is one of the primary ways in which SiMLR may be used
to study high-dimensional datasets.

The main metric will be the R$^2$ value between the model fit and the true
latent signal in the test data.  Higher values (averaged over simulations)
are associated with better performance.

The simulation study has several parameter the user may explore to gain more
insight. Default options corrupt half of the 3rd matrix which is why the sparse
SiMLR result outperforms SVD.  If we do not corrupt that data -- and use
joint SVD -- then performance is comparable.  If we do not corrupt data
and perform SVD separately on each matrix, then SiMLR somewhat outperforms
SVD, as expected.

Note: every run re-simulates the input data, re-runs SiMLR and SVD and compares
the findings to the same applied to permuted data.  In a real example, you would
only run SiMLR and SVD on permuted data in each loop.

- Result: Demonstrate well-above chance recovery of latent signal competitive
with SVD.  Also show that SiMLR performs better when there is only sparse signal
in even just one matrix.

```{r}
set.seed( 6 )
library(ANTsR)
compareToJointSVD = TRUE
doCorruption = TRUE
usePCA = FALSE
nSimlrEmbeddings = 6
nsims = 50 # number of simulation runs
if ( ! exists( "energyType" ) ) energyType = 'regression'
simlrRankEstimate <- function( targetVarx = 0.95, ... ) {
  # just run one permutation in order to get a baseline estimate
  j1 = scale( cbind( ... ), T, T )
  svd1 = svd( j1  )$d
  # how many do we need to explain XX% variance?
  varx = cumsum( svd1 ) / sum( svd1 )
  wv = min( which( varx >= targetVarx ) )
  j1p = scale( matrix( j1[ sample( prod( dim( j1 ) ) ) ], nrow=nrow(j1) ), T, T )
  svdperm = svd( j1p )$d
  varxp = cumsum( svdperm ) / sum( svdperm )
  myrankComp = svd1 > svdperm
  rm( j1, j1p )
  gc()
  return( list( svdD_original = svd1, svdD_perm = svdperm,
    varxComparison = myrankComp, suggestedRank = max( which(myrankComp[1:wv]) ) ) )
}
nsub = 100 # number of subjects
npix = round(c(200,100,500)/1)  # size of matrices
nk = nSimlrEmbeddings    # n components
mx = 'ica'   # mixing method
nits = 10  # n-iterations
nz = 0       # simulated signal parameter - not sensitive to this choice
nzs = 1.5    # simulated signal parameter - not sensitive to this choice
nzs2 = 1     # simulated signal parameter - not sensitive to this choice
train = c( rep(T,80) ,rep(F,20)) # train and test split
test = !train
nna = rep( NA, nsims )
simdatafrm = data.frame(
    symMeanCorrs = nna,
    symRSQ = nna,
    svdMeanCorrs = nna,
    svdRSQ = nna,
    prmMeanCorrs = nna,
    prmRSQ = nna
    )
for ( sim in 1:nsims ) {
    outcome = scale( matrix(runif( nsub * nk, nz, nzs2 ),ncol=nk) )
    view1tx = scale(matrix( rnorm( npix[1]  * nk, nz,nzs ), nrow=nk ))
    view2tx = scale(matrix( rnorm( npix[2]  * nk, nz,nzs ), nrow=nk ))
    view3tx = scale(matrix( rnorm( npix[3]  * nk, nz,nzs ), nrow=nk ))
    # here we mix the independent basis matrices with the true signal
    mixmats = diag( nk )
    outcomex=outcome
    reo=3:nk # throw some difference in here
    outcomex[,reo]=sample(outcomex[,reo])
    mat1 = (outcomex %*% mixmats[sample(1:nrow(mixmats)),] %*% view1tx)
    outcomex=outcome
    outcomex[,reo]=sample(outcomex[,reo])
    mat2 = (outcomex %*% mixmats[sample(1:nrow(mixmats)),] %*% view2tx)
    outcomex=outcome
    outcomex[,reo]=sample(outcomex[,reo])
    mat3 = (outcomex %*% mixmats[sample(1:nrow(mixmats)),] %*% view3tx)
    # add noise to each matrix
    mat1 = mat1 + matrix( rnorm( prod(dim(mat1)), 0, 0.25 ), nrow=nsub)
    mat2 = mat2 + matrix( rnorm( prod(dim(mat2)), 0, 0.25 ), nrow=nsub)
    mat3 = mat3 + matrix( rnorm( prod(dim(mat3)), 0, 0.25 ), nrow=nsub)

    if ( sim == 1 ) {
      myranker = simlrRankEstimate( 0.90, mat1, mat2, mat3 )
      myranker$suggestedRank
    }

    if ( doCorruption ) {
      # corrupt half of matrix 3
      mat3[ ,250:500] = matrix( rnorm( prod(dim(mat3[ ,250:500])), 10, 100 ), nrow=100)
    }

    r1 = cor( mat1 ) # regularization
    cthresh=0.66
    r1[ r1<cthresh] =  0
    r2 = cor( mat2 )
    r2[ r2<cthresh] =  0
    r3 = cor( mat3 )
    r3[ r3<cthresh] =  0
    result = simlr(
      list( vox = mat1[train,], vox2 = mat2[train,], vox3 = mat3[train,] ),
      smoothingMatrices = list( r1, r2, r3 ),
      energyType = energyType,
      initialUMatrix = nSimlrEmbeddings,
      verbose=T, iterations=nits, mixAlg=mx  )
    pred = predictSimlr( list(
      vox = mat1[train,], vox2 = mat2[train,], vox3 = mat3[train,] ), result )

    p1 = mat1 %*% abs(result$v[[1]])
    p2 = mat2 %*% abs(result$v[[2]])
    p3 = mat3 %*% abs(result$v[[3]])

    temp=data.frame( outc = outcome[,1], P1=p1[,1:2], P2=p2[,1:2], P3=p3[,1:2] )
    mdlsym=lm( outc~.,data=temp[train,])
    summary(mdlsym)
    rsqsym = cor( temp$outc[test], predict(mdlsym,newdata=temp[test,]) )^2

    # compare to permuted data
    s1 = sample( 1:nsub)
    s2 = sample( 1:nsub)
    s3 = sample( 1:nsub)
    pmat1=mat1[s1,]
    pmat2=mat2[s2,]
    pmat3=mat3[s3,]
    r1p = cor( pmat1 ) # regularization
    cthresh=0.66
    r1p[ r1p<cthresh] =  0
    r2p = cor( pmat2 )
    r2p[ r2p<cthresh] =  0
    r3p = cor( pmat3 )
    r3p[ r3p<cthresh] =  0
    resultp = simlr(list(
      vox = pmat1[train,], vox2 = pmat2[train,], vox3 = pmat3[train,] ),
      smoothingMatrices = list( r1p, r2p, r3p ),
      energyType = energyType,
      initialUMatrix = nSimlrEmbeddings,
      verbose = F, iterations=nits, mixAlg=mx  )
    predp = predictSimlr( list(
      vox = pmat1[train,], vox2 = pmat2[train,], vox3 = pmat3[train,] ), resultp )

#    print( sort(pred$aggregateTstats[1,], decreasing = T ) > sort(predp$aggregateTstats[1,], decreasing = T ) )
#    print( sort(pred$aggregateTstats[2,], decreasing = T ) > sort(predp$aggregateTstats[2,], decreasing = T ) )
#    print( sort(pred$aggregateTstats[3,], decreasing = T ) > sort(predp$aggregateTstats[3,], decreasing = T ) )

    p1p = pmat1 %*% abs(resultp$v[[1]])
    p2p = pmat2 %*% abs(resultp$v[[2]])
    p3p = pmat3 %*% abs(resultp$v[[3]])

    temp=data.frame( outc = outcome[,1], p1p[,1:2], p2p[,1:2], p3p[,1:2]   )
    mdlprm=lm( outc~.,data=temp[train,])
    rsqprm = cor( temp$outc[test], predict(mdlprm,newdata=temp[test,]) )^2

    # compare to SVD
    p1svd = mat1 %*% svd( mat1[train,], nu=0, nv=nk )$v
    p2svd = mat2 %*% svd( mat2[train,], nu=0, nv=nk )$v
    p3svd = mat3 %*% svd( mat3[train,], nu=0, nv=nk )$v
    if ( usePCA ) {
      p1svd = mat1 %*% prcomp( mat1[train,], rank.= nk )$rotation
      p2svd = mat2 %*% prcomp( mat2[train,], rank.= nk )$rotation
      p3svd = mat3 %*% prcomp( mat3[train,], rank.= nk )$rotation
    }
    temp=data.frame( outc = outcome[,1], p1svd[,1:2], p2svd[,1:2], p3svd[,1:2] )

    if ( compareToJointSVD ) {
        psvd = cbind( mat1,mat2,mat3) %*%
          svd( cbind( mat1,mat2,mat3), nu=0, nv=nk )$v
        if ( usePCA ) {
          psvd = cbind( mat1,mat2,mat3) %*%
            prcomp( cbind( mat1,mat2,mat3), rank.= nk  )$rotation
          }
        p1svd = psvd[,1:2]
        p2svd = psvd[,3:4]
        p3svd = psvd[,5:6]
        temp=data.frame( outc = outcome[,1], psvd[,1:6] )
    }
    mdlsvd=lm( outc~.,data=temp[train,])
    rsqsvd = cor( temp$outc[test], predict(mdlsvd,newdata=temp[test,]) )^2
simdatafrm[sim,] = c(
  mean(abs(cor( outcome, cbind(p1,p2,p3)))),
  rsqsym,
  mean(abs(cor( outcome, cbind(p1svd,p2svd,p3svd)))),
  rsqsvd,
  mean(abs(cor( outcome, cbind(p1p,p2p,p3p)))),
  rsqprm
  )
  print( simdatafrm[sim,] )

# plot( ts( colMeans( pred$aggregateTstats )[1:5]) )
  if ( sim == 1 & FALSE ) {
    resultFull = result
    nv = ncol( resultFull$v[[1]] )
    varxsums = c()
    for ( maxk in 2:nv ) {
      result = resultFull
      inmats = list(
        vox = mat1[train,], vox2 = mat2[train,], vox3 = mat3[train,] )
      for ( k in 1:length( inmats ) ) {
        result$u[[k]] = inmats[[k]] %*% result$v[[k]][,1:maxk]
        }
      pred = predictSimlr( inmats, result )
      pred$uOrder
      # reorder simlr output
      for ( k in 1:length( inmats ) ) {
        result$v[[k]] = result$v[[k]][,pred$uOrder]
        result$u[[k]] = inmats[[k]] %*% result$v[[k]][,1:maxk]
      }
      pred2 = predictSimlr( inmats, result )
      print( maxk )
      print( pred$varx )
      print( pred2$varx )
      varxsums[maxk-1] = sum( pred2$varx )
      }
  }
}

nms=c("symMeanCorrs","svdMeanCorrs","prmMeanCorrs")
h1 = hist( simdatafrm[,nms[1]])
h2 = hist( simdatafrm[,nms[2]])
h3 = hist( simdatafrm[,nms[3]])
myxl=c(0,0.7)
# pdf("./perfHistMeanCorr.pdf",width=6,height=4)
plot( h1, col=rgb(0,0,1,1/4), xlim=myxl, main='Simulation data: SyMILR vs SVD vs SyMILR-perm' ,  xlab='Mean Correlation' )
plot( h2, col=rgb(1,0,0,1/4), xlim=myxl, add=T )
plot( h3, col=rgb(0,1,0,1/4), xlim=myxl, add=T )
# dev.off()

nms=c("symRSQ","svdRSQ","prmRSQ")
h1 = hist( simdatafrm[,nms[1]])
h2 = hist( simdatafrm[,nms[2]])
h3 = hist( simdatafrm[,nms[3]])
myxl=c(0,1)
# pdf("./perfHistRSQ.pdf",width=6,height=4)
plot( h1, col=rgb(0,0,1,1/4), xlim=myxl, main='Simulation data: SyMILR (blu) vs SVD (grn) vs SyMILR-perm' ,  xlab='RSQ' )
plot( h2, col=rgb(1,0,0,1/4), xlim=myxl, add=T )
plot( h3, col=rgb(0,1,0,1/4), xlim=myxl, add=T )
```


## Look over results, statistically

The r-squared value is most informative as it tells us how well the omnibus
model predicts the known latent signal.

Compare SiMLR R$^2$ vs. SVD R$^2$ with paired t-test.

```{r}
print(t.test(simdatafrm[,"symRSQ"],simdatafrm[,"svdRSQ"],paired=T))
```


Compare SiMLR R$^2$ vs. permuted SiMLR R$^2$ with paired t-test.

```{r}
print(t.test(simdatafrm[,"symRSQ"],simdatafrm[,"prmRSQ"],paired=T))
```

Compare SVD R$^2$ vs. permuted SiMLR R$^2$ with paired t-test.

```{r}
print(t.test(simdatafrm[,"svdRSQ"],simdatafrm[,"prmRSQ"],paired=T))
```
