---
title: 'The Pediatric Template of Brain Perfusion: SyMLR'
author: "Brian B. Avants et al."
date: "`r Sys.Date()`"
output: html_document
---

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
bd="./data/"
set.seed(11)
library(ggplot2)
```

## Basic Setup

```{r dataio}
library(ANTsR)
library(visreg)
neanatvecs=2
thkmask=antsImageRead( paste(bd,"ptbp_mask_thickness.nii.gz",sep='') )
qth=0.05
mth='BH'
demog=read.csv(paste(bd,"ptbp_summary_demographics.csv",sep='') )
demog=demog[ , 1:19 ]
```

## Set up the formula

```{r form}
formulathkbase=" AgeAtScan * Sex + I(AgeAtScan^2)"
formulathkbase=" PIQ + BV"
formulathktest=" AgeAtScan * Sex + I(AgeAtScan^2)"
voi="AgeAtScan"
```

## Generic version: Function

```{r genericvox}
voxandeanatstudy <- function( demog, imgmat, imgmask,
                              formulabase, formulatest,
                              voi,
                              exclusionThresh,
                              baddata,
                              outprefix,
                              nv )
  {
  imgrowmeans=rowMeans( imgmat )
  wp = rep( FALSE, nrow( demog ) )
  for ( sub in unique( demog$SubID ) ) # get unique subjects
    {
    ww=which( demog$SubID == sub )
    ww=ww[ !is.na( imgrowmeans[ww] )  &
             imgrowmeans[ww] > exclusionThresh ]
    if ( length( ww ) > 0 ) wp[ ww[ 1 ] ] = TRUE
  }
 return( wp )
#  if ( ! all( is.na(baddata) ) ) # FIXME
#   wp[  which(wp==TRUE)[ baddata ] ]=FALSE # bad data
  ilist=matrixToImages( imgmat[wp,], imgmask )
#  for ( i in 1:length(ilist) ) ilist[[i]]=smoothImage( ilist[[i]], 2.0 )
  mydf=data.frame(
                 Sex=demog[wp,]$Sex,
                 AgeAtScan=demog[wp,]$AgeAtScan,
                 PIQ=demog[wp,]$Performance.IQ,
                 VIQ=demog[wp,]$Verbal.IQ,
                 BV=demog[wp,]$BV,
                 Cortex=demog[wp,]$Cortex,
                 LadderCom=demog[wp,]$Teen.Ladder.Community.Score,
                 LadderSES=demog[wp,]$Teen.Ladder.SES.score,
                 Income=demog[wp,]$Income )
  for ( kk in 2:ncol(mydf) ) mydf[,kk]=antsrimpute( mydf[,kk] )
  imgmat=imageListToMatrix( ilist, imgmask )
  locform=formula( paste( "imgmat ~", formulabase ,"+", formulatest ) )
  mdl=lm( locform, data=mydf )
  voxlm=bigLMStats( mdl , 1.e-8 )
  print(paste("begin low",outprefix) )
  lowmat = scale( lowrankRowMatrix( imgmat , 10 ), scale=FALSE )
  print(paste("begin eanat",outprefix) )
  imgeanat=sparseDecom( lowmat , imgmask, nvecs=nv, smooth=0.0,
    sparseness=0.05, cthresh=20000, mycoption=1, its=1 )
  rm( lowmat )
  print(paste("end eanat",outprefix) )
  eproj=abs(  imgeanat$eig  )
  eproj=( eproj/rowSums(eproj) )
  imgproj = imgmat %*% t( eproj  )
  mydf=data.frame( mydf, imgproj )
  print( names(mydf) )
  formbase=formula( paste( "imgproj ~", formulabase ) )
  formtest=formula( paste( "imgproj ~", formulabase ,"+", formulatest ) )
  mdl1=lm( formbase, data=mydf )
  mdl2=lm( formtest, data=mydf )
  eanatlm=bigLMStats( mdl2 , 1.e-6 )
  eseg=eigSeg( mask=imgmask, matrixToImages( imgeanat$eig, imgmask), F  )
  ofn=paste( outprefix, '_eseg.nii.gz', sep='' )
  antsImageWrite( eseg , ofn )
  anv=anova( mdl1, mdl2 )
  return( list(mydf=mydf,
               voxlm=voxlm,
               eanatlm=eanatlm,
               eseg=eseg,
               imgproj=imgproj,
               whichSubjects=wp,
               eanat=imgeanat,
               anv=anv) )
  }
```


# Three little modalities


## Thickness Eanat

```{r thkfun}
############################################
thkmask=antsImageRead( paste(bd,"ptbp_mask_thickness.nii.gz",sep='') )
thkfn=paste(bd,"ptbp_vox_thk.mha",sep='')
thkmat=as.matrix( antsImageRead( thkfn ) )
thkmat[ is.na( thkmat ) ] = 0
baddata=NA
if ( ! exists("wthk")  ) {
  wthk=voxandeanatstudy( demog, thkmat, thkmask,
         formulathkbase, formulathktest, voi=voi,
         exclusionThresh = 0.5, baddata=baddata,
         outprefix='/tmp/Eanat_THK', nv=neanatvecs )
#  qv=p.adjust( ethk$eanatlm$beta.pval[ voi, ], method='BH' )
 # print( min(qv) )
  }
```

## FA Eanat

```{r fafun}
fafn=paste(bd,"ptbp_vox_fa.mha",sep='')
famask=antsImageRead( paste(bd,"ptbp_mask_fa.nii.gz",sep='') )
famat=as.matrix( antsImageRead( fafn ) )
famat[ is.na( famat ) ] = 0
baddata=c(42,44) # FA
if ( ! exists("wfa")  ) {
  wfa=voxandeanatstudy( demog, famat, famask,
         formulathkbase, formulathktest, voi=voi,
         exclusionThresh = 0.2, baddata=baddata,
         outprefix='/tmp/Eanat_FA', nv=neanatvecs )
 # qv=p.adjust( efa$eanatlm$beta.pval[ voi, ], method='BH' )
#  print( min(qv) )
  }
```

## CBF Eanat

```{r cbffun}
cbffn=paste(bd,"ptbp_vox_cbf.mha",sep='')
cbfmat=as.matrix( antsImageRead( cbffn ) )
cbfmat[ is.na( cbfmat ) ] = 0
baddata=NA
if ( ! exists("wcbf") ) {
  wcbf=voxandeanatstudy( demog, cbfmat, thkmask,
         formulathkbase, formulathktest, voi=voi,
         exclusionThresh = 45, baddata=baddata,
         outprefix='/tmp/Eanat_CBF', nv=neanatvecs )
#  qv=p.adjust( ecbf$eanatlm$beta.pval[ voi,], method='BH' )
 # print( min(qv) )
  }
```

# Three little modalities jointly

## SyMLR

```{r symlr}
haveAllMerge=( wthk & wfa & wcbf )
inmats = list( thkmat[haveAllMerge,], famat[haveAllMerge,], cbfmat[haveAllMerge,] )
svdth = svd(inmats[[1]],nu=20,nv=0)$u
svdcb = svd(inmats[[2]],nu=20,nv=0)$u
svdfa = svd(inmats[[3]],nu=20,nv=0)$u
for ( k in 1:length( inmats ) ) inmats[[k]] = inmats[[k]] / norm( inmats[[k]], "F" )
if ( ! exists( "smoms") ) { # regularization matrices
  smoms = list()
  spatmat = t( imageDomainToSpatialMatrix( thkmask, thkmask ) )
  smoms[[1]] = knnSmoothingMatrix( spatmat, k = 100, sigma = 20 )
  spatmat = t( imageDomainToSpatialMatrix( famask, famask ) )
  smoms[[2]] = knnSmoothingMatrix( spatmat, k = 100, sigma = 20 )
  smoms[[3]] = smoms[[1]]
}
if ( ! exists("initu" ) ) initu = initializeSyMLR( inmats, 8, jointReduction = T, uAlgorithm = 'random' )
if ( ! exists("inmatsrob")) {
  inmatsrob=list()
  for ( k in 1:length( inmats ) ) inmatsrob[[k]] = robustMatrixTransform( inmats[[k]] )
}
if ( ! exists( "mysym" ) ) {
  mysym = symlr( voxmats = inmats, smoothingMatrices = smoms,
      iterations = 12,
      sparsenessQuantiles = c(0.5,0.5,0.5),
      positivities = rep( "positive", length( inmats )),
      lineSearchRange = c( -1, 1 )*10,
      lineSearchTolerance = 1e-4,
      mixAlg = 'pca', randomSeed = as.numeric(Sys.time()),
      initialUMatrix = 10,  energyType='cca',
      orthogonalize = F,  verbose = 1, constraint='none' )
  }
pthk = inmats[[1]] %*% ( mysym$v[[1]] )
pfa = inmats[[2]] %*% ( mysym$v[[2]] )
pcbf = inmats[[3]] %*% ( mysym$v[[3]] )
nmm=c("thk","fa","pcbf")
######
ss=1:6
svdlr = svd( cbind(pthk,pfa,pcbf)  )$u
m1 = lm( AgeAtScan ~   Sex, data = demog[haveAllMerge,] )
m2 = lm( AgeAtScan ~  ( pthk[,ss] + pcbf[,ss] + pfa[,ss] ) + Sex, data = demog[haveAllMerge,] )
#m2 = lm( AgeAtScan ~  svdlr + Sex, data = demog[haveAllMerge,] )
print( anova( m1, m2  ) )
print(mean(abs( demog[haveAllMerge,"AgeAtScan"] - predict(m2))))
m3 = lm( AgeAtScan ~  svdth + svdcb + svdfa + Sex, data = demog[haveAllMerge,] )
print( anova( m1, m3  ) )
mean(abs( demog[haveAllMerge,"AgeAtScan"] - predict(m3)))
# booyah!
m1 = lm( Verbal.IQ ~   Sex + AgeAtScan , data = demog[haveAllMerge,] )
m2 = lm( Verbal.IQ ~  ( pthk[,ss] +  pfa[,ss]  +pcbf[,ss]) + Sex +  AgeAtScan , data = demog[haveAllMerge,] )
print( anova( m1, m2  ) )
demog[haveAllMerge,"Income"] = antsrimpute(demog[haveAllMerge,"Income"])
m1 = lm( Teen.Ladder.SES.score ~   Sex + stats::poly(AgeAtScan,1), data = demog[haveAllMerge,] )
m2 = lm( Teen.Ladder.SES.score ~  ( pthk[,ss] + pfa[,ss]+ pcbf[,ss] ) + Sex + stats::poly(AgeAtScan,1)+1, data = demog[haveAllMerge,] )
# m2 = lm( Teen.Ladder.SES.score ~  ( svdth[,ss] + svdfa[,ss]+ svdcb[,ss] ) + Sex + AgeAtScan, data = demog[haveAllMerge,] )
print( anova( m1, m2  ) )
m1 = lm( FullScaleIQ ~   Sex + AgeAtScan, data = demog[haveAllMerge,] )
m2 = lm( FullScaleIQ ~  ( pthk[,ss] + pfa[,ss]+ pcbf[,ss] ) + Sex + AgeAtScan, data = demog[haveAllMerge,] )
#m2 = lm( FullScaleIQ ~  ( svdth[,ss] + svdcb[,ss]+ svdfa[,ss] ) + Sex + AgeAtScan, data = demog[haveAllMerge,] )
print( anova( m1, m2  ) )
# print( summary(m1) )
# print( summary(m2) )

print( summary(lm( pthk[,1] ~  ( pcbf[,2] + pfa[,2] ) + Sex + BV +
  AgeAtScan, data = demog[haveAllMerge,] )) )
# plot( thkmask, makeImage( thkmask,  abs(mysym$v[[1]][,1]) ) )
```


## Joint stats: Thickness, FA $+$ CBF

```{r jointstats,echo=FALSE}
alldf = data.frame( demog[haveAllMerge,], thk=pthk, fa=pfa, cbf=pcbf )
alldf = alldf[,c(3,4,20:49)]
mdl1=lm( AgeAtScan ~ . ,
  data=alldf[,c(1,2)])
mdl2=lm( AgeAtScan ~ . ,
  data=alldf[,])
knitr::kable(anova(mdl1,mdl2))
```


## Do some `visreg` ?

...

## Let's do some prediction

Use random forests to decide how relevant each
modality is to reproducible predictions of age.

First get the data organized.
```{r orgpred}
library(randomForest)
```



## Build the RF model


```{r rfpred,echo=FALSE}
groups <- rep( c(1,2), 1000 )[1:nrow(alldf)]
traing=groups==1
testg=groups==2
mdl=randomForest(  AgeAtScan ~ . , data=alldf[traing,], ntrees=2000 )
predage = predict( mdl, newdata=alldf[ testg, ] )
print( mean( abs(predage - alldf[testg,]$AgeAtScan ) ) )
temp=data.frame(predage=predage,realage=alldf[testg,]$AgeAtScan )
vmdl=lm( predage~realage,data=temp)
visreg::visreg(vmdl)
```

## RF importance plot

```{r rfimp,echo=FALSE}
impdf=data.frame(
  predNames=colnames(alldf)[ rev(order(mdl$importance)) ],
  importance=mdl$importance[ rev(order(mdl$importance)) ]
  )
knitr::kable(  impdf )
```




# SCCAN with ANTsR

## A little bit of CCA ...

First, FA-based prediction ...

* Use a couple of tricks

* Imputation

```{r facca,eval=TRUE}
sdemog=antsrimpute( demog[haveAllMerge,c(7,8)] )
# above is a bug sparseDecom/2 does not handle NAs well
````

* Residualization ...

```{r facca0,eval=TRUE}
rmat=residuals(lm( famat[haveAllMerge,] ~ demog$AgeAtScan[haveAllMerge] ))
```

* Why do we use these steps?

## A little bit of CCA ...

Now, run the decomposition - a "supervised PCA".

```{r facca2,eval=TRUE}
temp=rmat[traing,]
inmatrix=list( scale(temp) ,
 scale( data.matrix( sdemog[traing,])  ) )
if ( ! exists("sccan") )
  sccan=sparseDecom2(inmatrix, c(famask, NA),
     sparseness = c( 0.01, 0.3 ), verbose=0,
     nvecs = 3, its = 10, cthresh = c(5000, 0),
     perms = 0, smooth = 0.5, mycoption = 1 )
```

```{r faccalow,eval=FALSE}
# need this direction due to dimensionality restrictions
cog=antsrimpute( data.matrix( demog[haveAllMerge,c(3,6:12)]) )
dr=randcca( cog[traing,],  rmat[traing,], 6, 1 )
K = t(dr$svd$u)
lowmat = ( K %*% t(cog[traing,]) ) %*% rmat[traing, ]
if ( ! exists("sccan") )
  sccan=sparseDecom( lowmat, famask, nvecs=2, smooth=0.0,
    sparseness=0.05, cthresh=20000,  its=2 )
```


## Check the CCA regions

```{r thkcca4,eval=TRUE,echo=FALSE}
ccaeseg=eigSeg( famask, matrixToImages(t(sccan$eig1),famask), FALSE  )
plot( famask, ccaeseg, axis=3, nslices=16 )
```

## Look over cca results

Which demographic variables match which imaging variables?

```{r thkcca2,eval=TRUE}
rownames( sccan$eig2 )=colnames( sdemog )
knitr::kable( sccan$eig2 )
```

## Quick CCA-based prediction

PIQ prediction ...

```{r thkcca3,eval=TRUE,echo=FALSE}
sccanmat=sccan$eig1
sccanproj = famat[haveAllMerge,] %*% ( sccanmat )
ccadf=data.frame( sdemog, sccanproj[,-1], # look at mdl
  gender=demog$Sex[haveAllMerge],
  age=demog$AgeAtScan[haveAllMerge])
mdl=lm( Performance.IQ ~ . ,
  data=ccadf[traing,-1] )
pviq = predict( mdl, newdata=ccadf[ testg,-1] )
mycog=ccadf[testg,]$Performance.IQ
print( mean( abs( pviq - mycog) ) )
cor.test( pviq, mycog )
```


## More cca ...

Second, FA and CBF ...

These modalities relate strongly to age ... how do they covary?

```{r faccacbf,eval=TRUE}
sdemog=antsrimpute( demog[haveAllMerge,c(3,7,8)] )
temp=famat[haveAllMerge,][traing,]
for ( i in 1:nrow(temp) )
  temp[i,]=temp[i,]-mean(temp[i,])
temp2=cbfmat[haveAllMerge,][traing,]
for ( i in 1:nrow(temp2) )
    temp2[i,]=temp2[i,]-mean(temp2[i,])
inmatrix=list( temp , temp2 )
if ( ! exists("fcsccan") )
  fcsccan=sparseDecom2(inmatrix, c(famask, thkmask),
     sparseness = c( 0.05, -0.05 ),
     nvecs = 3, its = 10, cthresh = c(5000, 5000),
     perms = 0, smooth = 0.5, mycoption = 1 )
```


```{r faccacbflow,eval=FALSE}
temp=scale( famat[haveAllMerge,][traing,] )
for ( i in 1:nrow(temp) )
  temp[i,]=temp[i,]-mean(temp[i,])
temp2=scale( cbfmat[haveAllMerge,][traing,] )
# for ( i in 1:nrow(temp2) )
#  temp2[i,]=temp2[i,]-mean(temp2[i,])
ranker=20
dr=randcca( temp2,  temp, ranker, 1 )
K = t(dr$svd$u)
lowmat = ( K %*% t(temp2) ) %*% temp
s <- svd( temp, nu = 0, nv = ranker )$v
s2 <- svd( temp2, nu = 0, nv = ranker )$v
innermat <- t( temp2 %*% s2 ) %*% ( temp %*% s )
# if ( ! exists("fcsccan") )
#  fcsccan=sparseDecom(  lowmat, famask, cthresh = 5000,
#     sparseness = 0.05, nvecs = 2, its = 2, smooth = 0.0 )
if ( ! exists("ccsccan") ) {
  dr=randcca( temp,  temp2, ranker, 1 )
  K2 = t(dr$svd$u)
  lowmat2 =  ( K2 %*% t(temp) ) %*% temp2
#  ccsccan=sparseDecom( lowmat2, thkmask, cthresh = 5000,
#     sparseness = 0.05, nvecs = 2, its = 2, smooth = 0. )
  }

inmatrix=list( lowmat , lowmat2 )
fcsccan=sparseDecom2(inmatrix, c(famask, thkmask),
     sparseness = c( 0.05, -0.05 ),
     nvecs = 3, its = 10, cthresh = c(5000, 5000),
     perms = 0, smooth = 0.5, mycoption = 1 )
fcsccan$projections[,1]
```

## Check the CCA regions:FA

```{r faccavfa,eval=TRUE,echo=FALSE}
# ccaeseg=eigSeg( famask, matrixToImages(t(sccan$eig1),famask), FALSE  )
# ccaeseg=eigSeg( famask, fcsccan$eigenanatomyimages, FALSE  )
# plot( famask, ccaeseg, axis=3, nslices=16 )
```

## Check the CCA regions:CBF

```{r faccavcbf,eval=FALSE,echo=FALSE}
# ccaeseg=eigSeg( thkmask, ccsccan$eig, FALSE  )
# plot( thkmask, ccaeseg, axis=3, nslices=16 )
```

## Quick CCA-based prediction: FA and CBF

```{r facca3cbf,eval=TRUE,echo=FALSE}
fcsccanmat=fcsccan$eig1
# fcsccanmat=fcsccanmat/rowSums(fcsccanmat)
fcsccanproj = famat[haveAllMerge,] %*% ( fcsccanmat )
fcsccanmat2=fcsccan$eig2
# fcsccanmat2=fcsccanmat2/rowSums(fcsccanmat2)
fcsccanproj2 = cbfmat[haveAllMerge,] %*% ( fcsccanmat2 )
ccadf=data.frame( sdemog, fcsccanproj, fcsccanproj2,
  gender=demog$Sex[haveAllMerge] )
lmdl=lm( AgeAtScan ~ ., data=ccadf[traing,c(1,4:7)] )
mdl=randomForest( AgeAtScan ~., data=ccadf[traing,c(1,4:7)] )
pviq = predict( lmdl, newdata=ccadf[ testg,c(1,4:7) ] )
print( mean( abs( pviq - ccadf[testg,]$AgeAtScan) ) )
cor.test( pviq, ccadf[testg,]$AgeAtScan )
cor( fcsccanproj , fcsccanproj2 )
```





# Review

## Discussion

* Reviewed sparse dimensionality reduction

* These methods update traditional dimensionality reduction tools for neuroimaging

* We showed strategies for using these modalities together to find meaningful predictors.

* Performed some visualization and interpretation.
